---
title: "Analysis of Effects of Types of Storms and Other Severe Weather Events"
author: "Thammanoon Kawinfruangfukul"
date: "Friday, November 21, 2014"
output: html_document
---

## Synopsis
This analysis aims to identify types of storms and other severe weather events that affect to population health and economic consequences. The data comes from the National Weather Service Storm Data. Event types and other factors are normalized in order to specify top-10 event types affecting population health or economic consequences. 

## Data Processing
This section is divided into two parts: obtaining and cleaning data. In Obtaining data part, the raw will be downloaded from the URL referred to the National Weather Service Storm Data and loaded into r object called rawData.

```{r echo=TRUE, cache=TRUE}
# Download raw data and record the download date and time
if ( !file.exists( "data" ) ) {
    dir.create( "data" )
    }

url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "./data/stormData.csv.bz2"
download.file(url=url, destfile=destfile)
downloadOn <- format(Sys.time(), "%b %d, %Y %X")

# Load data
rawData <- read.csv(file =bzfile(destfile), header=TRUE)
```

Raw data is downloaded from `r url` on `r downloadOn`.  

In cleaning data part, the rawData will be subseted and normalized. Property and crop damages will be converted into dollar unit by multipling its value and unit. Morever, typos of Event types will be corrected.

```{r echo=TRUE, cache=TRUE}
# Subset data set
data <- rawData[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
colnames(data)[1:3] <- c("evtype", "fatalities", "injuries")

# Convert Damages into dollar
# Valid units are in alphabetical characters
multiplyConvertor <- function(value, unit) {
    multiply <- 1
    
    if(unit == "h" || unit == "H") {
        multiply <- 100
    }
    else if(unit == "k" || unit == "K") {
        multiply <- 1000
    }
    else if(unit == "m" || unit == "M") {
        multiply <- 1000000
    }
    else if(unit == "b" || unit == "B") {
        multiply <- 1000000000
    }
    
    return(value*multiply)
}

data$propdmg <- multiplyConvertor(data$PROPDMG, data$PROPDMGEXP)
data$cropdmg <- multiplyConvertor(data$CROPDMG, data$CROPDMGEXP)

# Remove unused columns
data$PROPDMG <- NULL
data$PROPDMGEXP <- NULL
data$CROPDMG <- NULL
data$CROPDMGEXP <- NULL

# Normalize event types
data$evtype[grep("avalanche", data$evtype, ignore.case = TRUE)] <- "AVALANCHE"
data$evtype[grep("THUNDERSTORM", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("THUDERSTORM", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("THUNDERESTORM", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("THUNDERSTROM", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("THUNDEERSTORM", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("TSTM", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("THUNERSTORM", data$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
data$evtype[grep("FLOOD", data$evtype, ignore.case = TRUE)] <- "FLOOD"
data$evtype[grep("TORNADO", data$evtype, ignore.case = TRUE)] <- "TORNADO"
data$evtype[grep("TORNDAO", data$evtype, ignore.case = TRUE)] <- "TORNADO"
data$evtype[grep("HAIL", data$evtype, ignore.case = TRUE)] <- "HAIL"
data$evtype[grep("HURRICANE", data$evtype, ignore.case = TRUE)] <- "HURRICANE"
data$evtype[grep("HEAT", data$evtype, ignore.case = TRUE)] <- "HEAT"
data$evtype[grep("FREEZ", data$evtype, ignore.case = TRUE)] <- "FREEZE"
data$evtype[grep("Frost", data$evtype, ignore.case = TRUE)] <- "FREEZE"
data$evtype[grep("dust", data$evtype, ignore.case = TRUE)] <- "DUST STORM"
data$evtype[grep("slide", data$evtype, ignore.case = TRUE)] <- "LANDSLIDE"
data$evtype[grep("wind", data$evtype, ignore.case = TRUE)] <- "HIGH WIND"
data$evtype[grep("wnd", data$evtype, ignore.case = TRUE)] <- "HIGH WIND"
data$evtype[grep("snow", data$evtype, ignore.case = TRUE)] <- "SNOW"
data$evtype[grep("COLD", data$evtype, ignore.case = TRUE)] <- "COLD"
data$evtype[grep("ice", data$evtype, ignore.case = TRUE)] <- "ICE"
data$evtype[grep("icy", data$evtype, ignore.case = TRUE)] <- "ICE"
data$evtype[grep("LIGHT", data$evtype, ignore.case = TRUE)] <- "LIGHTNING"
data$evtype[grep("Volca", data$evtype, ignore.case = TRUE)] <- "VOLCANIC ERUPTION"
data$evtype[grep("dry", data$evtype, ignore.case = TRUE)] <- "DRY"
data$evtype[grep("WATERSPOUT", data$evtype, ignore.case = TRUE)] <- "WATERSPOUT"
data$evtype[grep("Rain", data$evtype, ignore.case = TRUE)] <- "RAIN"
```

## Results
Effects to population health should be considered both fatalities and injuries. Since fatalities are more harmful to population health, Simple additive weighting method [SAW] is adopted. Each factor will be associated with the assigned weight.
```{r echo=TRUE, cache=TRUE, warning=FALSE}
health <- aggregate(data$fatalities*0.7+data$injuries*0.3,
                     by=list(data$evtype),
                     FUN=sum)
colnames(health) <- c("evtype", "totalHealth")
rankedHealth <- health[order(health$totalHealth, decreasing=TRUE), ]

library(ggplot2)
g <- ggplot(rankedHealth[1:10, ], aes(x = reorder(evtype, totalHealth), 
                                      y = totalHealth))
g <- g + geom_bar(stat="identity")
g <- g + theme(axis.text.x = element_text(angle = 90, hjust = 1))
g <- g + labs(x = "Event type")
g <- g + labs(y = "Weighted effect to health population")
g <- g + labs(title = "Fig. 1 Top-10 Event Types Affected to Health Population")
print(g)
```

Effects to economic consequences should be considered both property and crop damages equally since both damages are relatively affected economic consequnces.
```{r echo=TRUE, cache=TRUE, warning=FALSE}
damages <- aggregate(data$propdmg+data$cropdmg,
                     by=list(data$evtype),
                     FUN=sum)
colnames(damages) <- c("evtype", "totalDamage")
rankedDamages <- damages[order(damages$totalDamage, decreasing=TRUE), ]

library(ggplot2)
g <- ggplot(rankedDamages[1:10, ], aes(x = reorder(evtype, totalDamage), 
                                       y = totalDamage))
g <- g + geom_bar(stat="identity")
g <- g + theme(axis.text.x = element_text(angle = 90, hjust = 1))
g <- g + labs(x = "Event type")
g <- g + labs(y = "Total damages ($)")
g <- g + labs(title = "Fig. 2 Top-10 Event Types Affected to Economic Consequences")
print(g)
```

Based on figure 1 and 2, tornado is the most harmful with respect to population health and has the greatest economic consquences. Flood, thunderstorm, lightning are considered important event types since they are on top-5 of both effects to population health and economic consequences.

## References
* [SAW] Yoon, K. P., & Hwang, C.-L. (1995). Multiple Attribute Decision Making: An Introduction. Thousand Oaks, CA: Sage Publications
